---
title: "TimeScape vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TimeScape vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

TimeScape is a visualization tool for temporal clonal evolution.

# Installation 

To install TimeScape, type the following commands in R:

```{r, eval=FALSE}
# try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("timescape")
```

# Examples 

Run the examples by: 

```{r, eval=FALSE}
example("timescape")
```

The following visualizations will appear in your browser (optimized for Chrome):

The first visualization is of the acute myeloid leukemia patient from Ding et al., 2012.
The second visualization is of the metastatic ovarian cancer patient 7 from McPherson and Roth et al., 2016.

```{r, echo=FALSE}
library(devtools)  
#install_bitbucket("shahlabbcca/timescape")  
library(timescape)  
```
```{r, echo=FALSE}
# EXAMPLE 1 - Acute myeloid leukemia patient, Ding et al., 2012
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "AML_tree_edges.csv", package = "timescape"))
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "AML_clonal_prev.csv", package = "timescape"))
# targeted mutations
mutations <- read.csv(system.file("extdata", "AML_mutations.csv", package = "timescape"))
# perturbations
perturbations <- data.frame( pert_name = c("Chemotherapy"), 
                             prev_tp = c("Diagnosis"),
                              frac = c(0.1))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, perturbations = perturbations)
```

<br>
<br>
<br>
<br>
<br>

```{r, echo=FALSE}
# EXAMPLE 2 - Patient 7, McPherson and Roth et al., 2016
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "px7_tree_edges.csv", package = "timescape"))
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "px7_clonal_prev.csv", package = "timescape"))
# clone colours
clone_colours <- data.frame(clone_id = c("A","B","C","D","E"), 
                            colour = c("d0ced0", "2CD0AB", "FFD94B", "FD8EE5", "F8766D"))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, clone_colours = clone_colours, height=260, alpha=15)
```

# Required parameters 

The required parameters for TimeScape is structure as follows:

$clonal\_prev$ is a data frame consisting of clonal prevalences for each clone at each time point. The columns in this data frame are:

1. character() $timepoint$ - time point
2. character() $clone\_id$ - clone id
3. numeric() $clonal\_prev$ - clonal prevalence.

$tree\_edges$ is a data frame describing the edges of a rooted clonal phylogeny. The columns in this data frame are:

1. character() $source$ - source node id
2. character() $target$ - target node id.

$mutations$ is a data frame consisting of the mutations originating in each clone. The columns in this data frame are:

1. character() $chrom$ - chromosome number
2. numeric() $coord$ - coordinate of mutation on chromosome
3. character() $clone\_id$ - clone id
4. character() $timepoint$ - time point
5. numeric() $VAF$ - variant allele frequency of the mutation in the corresponding timepoint. 

# Optional parameters 

## Clone colours 

Clone colours may be changed using the $clone\_colours$ parameter, for instance, compare the default colours :

```{r, echo=FALSE}

# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=15)
```

with specified custom colours:

```{r, echo=FALSE}
# clone colours
clone_colours <- data.frame(clone_id = c("A","B","C","D","E"), 
                            colour = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00"))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, clone_colours = clone_colours, height=260, alpha=15)
```

## Alpha value

The alpha value of each colour may be tweaked in the $alpha$ parameter (a numeric value between [0, 100]). Compare alpha of 10: 
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=10)
```

with the alpha value of 90:
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=90)
```

## Titles 

The x-axis, y-axis and phylogeny titles may be changed using the $xaxis\_title$, $yaxis\_title$ and $phylogeny\_title$ parameters, which take in a character string.

Here are some custom titles:
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=15, xaxis_title="My X Axis", yaxis_title="My Y Axis", phylogeny_title="My Phylogeny")
```

## Genotype position

The position of each genotype with respect to its ancestor can be altered. The “stack” layout is the default layout. It stacks genotypes one on top of another to clearly display genotype prevalences at each time point. The “space” layout uses the same stacking method while maintaining (where possible) a minimum amount of space between each genotype. The “centre” layout centers genotypes with respect to their ancestors. Here we'll see an example of each:

### Stacked
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=15, genotype_position="stack")
```

### Spaced
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=15, genotype_position="space")
```

### Centered
```{r, echo=FALSE}
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, height=260, alpha=15, genotype_position="centre")
```

## Perturbations

Perturbations may be added to the TimeScape using the $perturbations$ parameter, which is a data frame consisting of the following columns:

1. character() $pert_name$ - the perturbation name
2. character() $prev_tp$ - the time point (as labelled in clonal prevalence data) BEFORE perturbation
3. numeric() $frac$ - the fraction of total tumour content remaining at the time of perturbation, range [0, 1].

### Without perturbation

```{r, echo=FALSE}
# EXAMPLE 1 - Acute myeloid leukemia patient, Ding et al., 2012
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "AML_tree_edges.csv", package = "timescape"))
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "AML_clonal_prev.csv", package = "timescape"))
# targeted mutations
mutations <- read.csv(system.file("extdata", "AML_mutations.csv", package = "timescape"))

# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges)
```

### With perturbation (reduces the tumour to 10% of its original size)

```{r, echo=FALSE}
# perturbations
perturbations <- data.frame( pert_name = c("Chemotherapy"), 
                             prev_tp = c("Diagnosis"),
                              frac = c(0.1))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, perturbations = perturbations)
```

# Documentation 

To view the documentation for TimeScape, type the following command in R:

```{r, eval=FALSE}
?timescape
```

TimeScape was developed at the Shah Lab for Computational Cancer Biology at the BC Cancer Research Centre.

References:

Ding, Li, et al. "Clonal evolution in relapsed acute myeloid leukaemia revealed by whole-genome sequencing." Nature 481.7382 (2012): 506-510.

McPherson, Andrew, et al. "Divergent modes of clonal spread and intraperitoneal mixing in high-grade serous ovarian cancer." Nature genetics (2016).