---
title: "TimeScape vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TimeScape vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

TimeScape is a visualization tool for temporal clonal evolution.

To run TimeScape, type the following commands in R:

```{r, eval=FALSE}
install.packages("devtools") # if not already installed  
library(devtools)  
install_bitbucket("MO_BCCRC/timescape")  
library(timescape)  
example(timescape) # to run example
```

And the following visualizations will appear in your browser (optimized for Chrome):

The first visualization is of the acute myeloid leukemia patient from Ding et al., 2012.
The second visualization is of the metastatic ovarian cancer patient 7 from McPherson and Roth et al., 2016.

<br>
<br>

```{r, echo=FALSE}
library(devtools)  
install_bitbucket("MO_BCCRC/timescape")  
library(timescape)  
```
```{r, echo=FALSE}
# EXAMPLE 1 - Acute myeloid leukemia patient, Ding et al., 2012
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "AML_tree_edges.csv", package = "timescape"))
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "AML_clonal_prev.csv", package = "timescape"))
# targeted mutations
mutations <- read.csv(system.file("extdata", "AML_mutations.csv", package = "timescape"))
# perturbations
perturbations <- data.frame( pert_name = c("Chemotherapy"), 
                             prev_tp = c("Diagnosis"),
                              frac = c(0.1))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, perturbations = perturbations)
```

<br>
<br>
<br>
<br>
<br>

```{r, echo=FALSE}
# EXAMPLE 2 - Patient 7, McPherson and Roth et al., 2016
# genotype tree edges
tree_edges <- read.csv(system.file("extdata", "px7_tree_edges.csv", package = "timescape"))
# clonal prevalences
clonal_prev <- read.csv(system.file("extdata", "px7_clonal_prev.csv", package = "timescape"))
# clone colours
clone_colours <- data.frame(clone_id = c("A","B","C","D","E"), 
                            colour = c("d0ced0", "2CD0AB", "FFD94B", "FD8EE5", "F8766D"))
# run timescape
timescape(clonal_prev = clonal_prev, tree_edges = tree_edges, clone_colours = clone_colours, height=260, alpha=15)
```


<br>
<br>
<br>
To view the documentation for TimeScape, type the following command in R:

```{r, eval=FALSE}
?timescape
```

TimeScape was developed at the Shah Lab for Computational Cancer Biology at the BC Cancer Research Centre.

References:

Ding, Li, et al. "Clonal evolution in relapsed acute myeloid leukaemia revealed by whole-genome sequencing." Nature 481.7382 (2012): 506-510.

McPherson, Andrew, et al. "Divergent modes of clonal spread and intraperitoneal mixing in high-grade serous ovarian cancer." Nature genetics (2016).